---
title: "Testing anova functions"
pagetitle: "aah | exp 1-2 | analysis"
author: "Owen Morgan"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: html_document
---

<style type="text/css">
  body{
  font-family: Avenir;
  font-size: 12pt;
}
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.align = "center", fig.width = 9,
                      fig.height = 6, results = "markup")
options(knitr.kable.NA = "")

cli.progress_show_after <- 0

## Do not use scientific notation until 9 decimal places.
options(scipen = 9, digits = 9)

# Suppress summarise info
options(dplyr.summarise.inform = FALSE)
```

```{r}
library(car)
library(palmerpenguins)
library(tidyverse)
library(here)
library(broom)
library(lme4)
## Function to make pretty table that rounds to 2 digits and stays in place
library(gt)
pretty_table <- function(table, title = NULL, digits = 3,
                         groupname_col = NULL
                         ) {
    gt(table, groupname_col = groupname_col) |> 
      tab_header(title = title) |> 
      sub_missing(columns = everything(), missing_text = "-") |> 
      fmt_number(columns = where(is.numeric),
                 drop_trailing_zeros = T,
                 decimals = digits) |> 
   tab_style(
      #Apply new style to all column headers
     locations = cells_column_labels(columns = everything()),
     style     = list(
       #Give a thick border below
       # cell_borders(sides = "bottom", weight = px(2)),
       #Make text bold
       cell_text(weight = "bold")
     )
   )
}


## Format p-values in a table
format_p.value <- function(tbl) {
  tbl  %>%
    ## display p as "<.001" if it is less than 0.001; or else, round.
    ## Diaply p as "<.0001 if less that .0001; or else, round.
    mutate(p.value = case_when(
      (p.value < 0.0001) ~ "<.0001",
      (0.0001 <= p.value & p.value < 0.001) ~
        as.character(p.value %>% round(4)),
      (0.001 <= p.value & p.value < 0.01) ~
        as.character(p.value %>% round(3)),
      (0.01 <= p.value) ~
        as.character(p.value  %>% round(3))
    )) |> 
    ## Remove leading zero
    mutate(p.value = p.value |> str_remove("^0+"))
}
```

# {.tabset}

## Notes

#### anova()
- "anova() is a generic function." ([source](https://stackoverflow.com/questions/40823310/when-should-i-use-aov-and-when-anova])) To find documentation, look at, e.g., ?anova.lm.
- "The standard R anova function calculates sequential ("type-I") tests. These rarely test interesting hypotheses in unbalanced designs." (?car::Anova)
- "Specifying a single object gives a sequential analysis of variance table for that fit. That is, the reductions in the residual sum of squares as each term of the formula is added in turn are given in as the rows of a table, plus the residual sum of squares. The table will contain F statistics (and P values) comparing the mean square for the row to the residual mean square. If more than one object is specified, the table has a row for the residual degrees of freedom and sum of squares for each model. For all but the first model, the change in degrees of freedom and sum of squares is also given. (This only make statistical sense if the models are nested.)" (?anova.lm)

#### aov()
-"aov is designed for balanced designs." (?aov)
-"The standard aov() function in base-R uses Type I sums of squares"([source](https://bookdown.org/ndphillips/YaRrr/type-i-type-ii-and-type-iii-anovas.html)

#### car::Anova()
- Can specify type 2 or 3 in the "type" argument. (?car::Anova)

#### Type I, II, and III sum-of-squares
- [stackexchange post](https://stats.stackexchange.com/questions/20452/how-to-interpret-type-i-type-ii-and-type-iii-anova-and-manova/20455#20455)
- [YaRrr](https://bookdown.org/ndphillips/YaRrr/type-i-type-ii-and-type-iii-anovas.html)



## Simple model (no random effects) {.tabset .tabset-pills}
```{r}
fava_e1 <- read_tsv("aah_long_e1.tsv")
fava_e2 <- read_tsv("aah_long_e2.tsv")

fava_raw <-  bind_rows(fava_e1, fava_e2)

fava <- fava_raw |>  filter(block_type == "main" & target_present == "yes" & correct == T
         & handedness %in% c("Left", "Right")) |> 
select(subject, field, level, handedness, rt)
```

```{r}
simple_model <- lm(data = fava, formula = rt ~ field * level * handedness)
```

### Type 1
anova(fm)
```{r}
anova(simple_model) |> tidy() |> format_p.value() |> pretty_table("anova(simple model)")
```
anova(fm, rm) [3-way] 
```{r}
reduced_model <- update(simple_model, . ~ . - field:level:handedness)
anova(simple_model, reduced_model) |> as_tibble() |>
  pretty_table("anova(fm, rm) [3-way]")
```
anova(fm, rm) [component 2-way] (Returns nonsense)
```{r}
reduced_model <- lm(data = fava, formula = rt ~ field + level + handedness + field:handedness + level:handedness + field:level:handedness)
anova(simple_model, reduced_model) |> as_tibble() |>
  pretty_table("anova(fm, rm) [2-way: field:level]")
```
anova(fm, rm) [component 1-way] (Returns nonsense)
```{r}
reduced_model <- lm(data = fava, formula = rt ~  level + handedness +  field:handedness + level:handedness + field:level + field:level:handedness)
anova(simple_model, reduced_model) |> as_tibble() |>
  pretty_table("anova(fm, rm) [1-way: field]")
```
aov(fm) (identical to anova(fm)
```{r}
aov(simple_model) |> tidy() |> format_p.value() |> pretty_table("aov(simple model)")
```
aov(fm, rm) [3-way]  (aov does not do this)

### Type 2
car::Anova()
```{r}
car::Anova(simple_model, type = 2) |> tidy() |> format_p.value() |> 
  pretty_table("car::Anova(simple_model, type 2)")
```
car::Anova(fm, rm) [3-way] (car does not do this)

### Type 3
car::Anova()
```{r}
car::Anova(simple_model, type = 3) |> tidy() |> format_p.value() |> 
  pretty_table("car::Anova(simple_model, type 3)")
```


## Mixed model (random intercept for penguin) {.tabset .tabset-pills}
```{r}
mixed_model <- lmer(data = fava, formula = rt ~ field * level * handedness + (1|subject))
```


### Type 1
anova(fm)
```{r}
anova(mixed_model) |> tidy() |> pretty_table("anova(mixed_model)")
```
anova(fm, rm) [3-way] 
```{r}
reduced_model <- update(mixed_model, . ~ . - field:level:handedness)
anova(mixed_model, reduced_model) |> as_tibble() |>
  pretty_table("anova(fm, rm) [3-way]")
```
anova(fm, rm) [component 2-way] (Returns nonsense)
```{r}
reduced_model <- lmer(data = fava, formula = rt ~ field + level + handedness + field:handedness + level:handedness + field:level:handedness + (1|subject))
anova(mixed_model, reduced_model) |> as_tibble() |>
  pretty_table("anova(fm, rm) [2-way: field:level]")
```
anova(fm, rm) [component 1-way] (Returns nonsense)
```{r}
reduced_model <- lmer(data = fava, formula = rt ~  level + handedness +  field:handedness + level:handedness + field:level + field:level:handedness + (1|subject))
anova(mixed_model, reduced_model) |> as_tibble() |>
  pretty_table("anova(fm, rm) [1-way: field]")
```
aov(fm) (identical to anova(fm)
aov does not run on our mixed model, with the error:
Error in fixed.only && random.only : invalid 'x' type in 'x && y'

### Type 2
car::Anova()
```{r}
car::Anova(mixed_model, type = 2) |> tidy() |> format_p.value() |> 
  pretty_table("car::Anova(mixed_model, type 2)")
```
car::Anova(fm, rm) [3-way] (car does not do this)

### Type 3
car::Anova()
```{r}
car::Anova(mixed_model, type = 3) |> tidy() |> format_p.value() |> 
  pretty_table("car::Anova(mixed_model, type 3)")
```



